A formal treatment of bidirectional typing

### 2023-10-13 tex/ESOP24/S1-intro.tex
[ tex ] Update Example section

Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism leads to undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
\varcitet{Dunfield2021}{ in their survey paper} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a term is also given so that there is more information to work with.
More formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is computed as output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can usually be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

@@

Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis; this situation is in stark contrast to parsing, which has general theories together with widely applicable techniques and practical tools, notably parser generators.
%\footnote{The same could be said for type checking in general, not just for bidirectional type systems.
%While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
While it is straightforward to derive a type synthesis algorithm,
for bidirectional typing, \citeauthor{Dunfield2021} only present informal design principles that the community learned from individual systems, rather than a general treatment providing logical specifications and rigorously proven properties for a class of systems.
Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.

To mitigate the above problem, we develop with the proof assistant \Agda a general and formal treatment of bidirectional typing and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a (scoped-checked) raw term as input.

@@

Our work adapts proof-relevant (bidirectional) type synthesis illustrated by \citet{Wadler2022}.
Their treatment deviates from the usual formulation: 
a type synthesis algorithm is traditionally presented as \emph{algorithmic rules} such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the surface language to produce $\isTerm{t}'$ of type $A$ in the core language.
Such an algorithm is then followed by soundness and completeness assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if sufficiently annotated.
By contrast, \citeauthor{Wadler2022} exploit the computational and logical nature of Martin-L\"of type theory and formulate \emph{algorithmic soundness, completeness, and decidability in one go}.
%(although they do not emphasise the difference of their approach from the traditional one)

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every $P$ in type theory, and we say $P$~is logically \emph{decidable} if the law holds for~$P$.
For example, a proof of \emph{decidable equality}, i.e. $\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly; such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
Further, suppose a proof of the following statement:
\begin{quote}
  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction'
\end{quote}
or rephrased succinctly as 
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
Computationally, the proof yields either a typing derivation for the given raw term~$t$ or a negation proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
Hence, both algorithmic soundness and completeness in original form are implied.
Hence, the formal proof of the statement as a program is a verified \emph{proof-relevant} type synthesiser, because it does not only return yes/no but also constructs a proof. 
%For a type synthesiser being proof-relevant, all of its intermediate programs have to be proof-relevant, including bidirectional type synthesiser and its mode decorator.

\Josh{emphasise the scope of our approach}
\subsubsection{Mode decoration and bidirectional type synthesis}
\Josh{A missing step between bidirectional type synthesis and parsing}
As raw terms may not be sufficiently annotated, we propose a preprocessing step called \emph{mode decoration} to determine whether a raw term can be processed by a bidirectional type synthesiser later.
A mode decorator constructs a mode derivation for a raw term or pinpoints missing annotations.
A bidirectional type synthesiser and a type synthesiser differ in their domain---bidirectional type synthesis only works for mode-decorated raw terms instead of all raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between \emph{bidirectional typing derivations} and \emph{typing derivations} with \emph{mode derivations} (of a raw term), which is a more refined and useful formulation than annotatability described by \citet[Section~3.2]{Dunfield2021}.
Our proof-relevant bidirectional type synthesiser is complete with respect to mode-decorated terms.

We also prove that a raw term with missing annotations is exactly a raw term without a mode derivation, so our mode decorator is indeed proof-relevant.

By combining mode-decoration and bidirectional type synthesis, we show a trichotomy on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation, suggesting that type synthesis should not be viewed as a bisection between typable and untypable raw terms but a trichotomy in general.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness,
  \item the concept of mode decoration and the comparison between our completeness and \varcitet{Dunfield2021}{'s annotatability}, 
    \Josh{generalisation mode decoration?}
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator for proof-relevant type-synthesisers.
\end{enumerate}
Our theory was first developed in \Agda and then translated to the mathematical vernacular, filling the gap between our informal presentation and its verified implementation.

>>>>

Type inference is an important mechanism for the transition to well typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism leads to undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
\varcitet{Dunfield2021}{ in their survey paper} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a term is also given so that there is more information to work with.
More formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is computed as output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can usually be `read off' from a well designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

@@

Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis; this situation is in stark contrast to parsing, which has general theories together with widely applicable techniques and practical tools, notably parser generators.
%\footnote{The same could be said for type checking in general, not just for bidirectional type systems.
%While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
While it is straightforward to derive a bidirectional type synthesis algorithm, \citeauthor{Dunfield2021} only present informal design principles that the community learned from individual systems, rather than a general treatment providing logical specifications and rigorously proven properties for a class of systems.
Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.

To mitigate the above problem, we develop with the proof assistant \Agda a general and formal treatment of bidirectional typing and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a raw term as input.

@@

Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for \PCF.
The proof-relevant formulation deviates from the usual one:
traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that annotations can be added to~$\isTerm{t}$ in the surface language%
\Josh{The algorithm adds annotations to the input term?}
to produce $\isTerm{t}'$ of type~$A$ in the core language.
Such an algorithm is then accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates \emph{algorithmic soundness, completeness, and decidability in one go}.
We explain this in a bit more detail below.
%(although they do not emphasise the difference of their approach from the traditional one)

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is also a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability implies algorithmic decidability.
%For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
%Such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
%In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
More specifically, consider the statement of the type inference problem
\begin{quote}
  `for a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction',
\end{quote}
which can be rephrased more succinctly as
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (and can be strengthened to the original form).
Hence, proving the statement is the same as constructing a verified proof-relevant type inferencer, which returns not only an answer but also a proof justifying that answer.

%Hence, both algorithmic soundness and completeness in original form are implied.
%For a type synthesiser being proof-relevant, all of its intermediate programs have to be proof-relevant, including bidirectional type synthesiser and its mode decorator.
%\Josh{emphasise the scope of our approach}

\subsubsection{Annotations in the type synthesis problem}
%\Josh{A missing step between bidirectional type synthesis and parsing}

As mentioned in the beginning, with bidirectional typing we avoid the generally undecidable problem of type inference, and instead solve the simpler problem about the typability of `sufficiently annotated' raw terms, which we call the type synthesis problem to distinguish it from type inference.
Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
In our theory, we define \emph{mode derivations} so that we can explicitly take annotations into account, in particular properly formulating the type synthesis problem.
We also propose a preprocessing step called \emph{mode decoration} that helps the user to work with annotations in practice.

In a bit more detail:
The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
To solve the type synthesis problem, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide whether the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provide information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
Then a bidirectional type synthesiser decides the typability of only mode-decorated raw terms instead of all raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more natural and useful than annotatability described by \citet[Section~3.2]{Dunfield2021}.

%As raw terms may not be sufficiently annotated, we propose a preprocessing step called \emph{mode decoration} to determine whether a raw term can be processed by a bidirectional type synthesiser later.
%A bidirectional type synthesiser and a type synthesiser differ in their domain---bidirectional type synthesis only works for mode-decorated raw terms instead of all raw terms.

%Our proof-relevant bidirectional type synthesiser is complete with respect to mode-decorated terms.
%We also prove that a raw term with missing annotations is exactly a raw term without a mode derivation, so our mode decorator is indeed proof-relevant.

%By combining mode-decoration and bidirectional type synthesis, we show a trichotomy on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation, suggesting that type synthesis should not be viewed as a bisection between typable and untypable raw terms but a trichotomy in general.
%\Josh{elaborate}

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness;
  \item mode derivations for explicitly dealing with annotations in the theory (in particular leading to a more natural and useful version of completeness than annotatability~\cite{Dunfield2021}), and mode decoration for helping the user to work with annotations in practice;
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator for proof-relevant type-synthesisers.
\end{enumerate}
Our theory was first developed in \Agda and then translated to the mathematical vernacular, filling the gap between our informal presentation and its verified implementation.

### 2023-10-13 tex/ESOP24/S2-outline.tex
[ tex ] Update Example section

\begin{theorem}\label{thm:implementation}
A type synthesiser that uses `not in synthesising mode' as its excuse can be constructed from a mode decorator and a bidirectional type synthesiser.
\end{theorem}

>>>>

\begin{theorem}\label{thm:implementation}
A type synthesiser using `not in synthesising mode' as its excuse can be constructed from a mode decorator and a bidirectional type synthesiser.
\end{theorem}
